Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-13 paper/preprint/paper.tex
rerendered full paper

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed Counterfactual Training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: Counterfactual Training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
  follow the standard ML convention, where ``degrees of freedom'' refer
  to the number of parameters estimated from data.} Indeed, recent work
on the so called ``lottery ticket hypothesis'' suggests that modern
neural networks can be pruned by up to 90\% while preserving their
predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
Carbin 2019}) and generalizability
(\citeproc{ref-morcos2019success}{Morcos et al. 2019}). Similarly, Zhang
et al. (\citeproc{ref-zhang2021understanding}{2021}) showed that
state-of-the-art neural networks are so expressive that they can fit
randomly labeled data. Thus, looking at the predictive performance
alone, the solutions may seem to provide compelling explanations for the
data, when in fact they are based on purely associative, semantically
meaningless patterns. This poses two related challenges. Firstly, there
is no dependable way to verify if such complex representations
correspond to meaningful and plausible explanations. Secondly, even if
we could resolve the first challenge, it remains undecided how to ensure
that models can \emph{only} learn valuable explanations.

@@

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data
(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
mitigates the aforementioned risk of relying on a single locally optimal
representations that corresponds to semantically meaningless
explanations for the data. Previous work by Schut et al.
(\citeproc{ref-schut2021generating}{2021}) similarly found that
generating plausible (``interpretable'') counterfactual explanations is
almost trivial for deep ensembles that have also undergone adversarial
training. The case for JEMs is even clearer: they involve a hybrid
objective that induces both high predictive performance and generative
capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
is closely related to the idea of aligning models with plausible
explanations and has inspired our proposed counterfactual training
objective, as we explain in Section~\ref{sec-method}.

@@

Counterfactual training (CT) combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit goal of aligning representations with plausible explanations
that comply with user requirements. In the context of CEs, plausibility
has broadly been defined as the degree to which counterfactuals comply
with the underlying data-generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CEs to provide
algorithmic recourse (AR) to individuals (negatively) affected by opaque
models. For AR recommendations to be actionable, they need to not only
result in plausible counterfactuals but also be attainable. A plausible
CE for a rejected 20-year-old loan applicant, for example, might reveal
that their application would have been accepted, if only they were 20
years older. Ignoring all other features, this would comply with the
definition of plausibility if 40-year-old individuals were in fact more
credit-worthy on average than young adults. But of course this CE does
not qualify for providing actionable recourse to the applicant since
\emph{age} is not a (directly) mutable feature. CT aims to improve model
explainability by aligning models with counterfactuals that meet both
desiderata: plausibility and actionability. Formally, we define
explainability as follows:

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation}

where \(\text{yloss}(\cdot)\) is a classification loss that induces
discriminative performance (e.g., cross-entropy). The second and third
terms in Equation~\ref{eq-obj} are explained in detail below. For now,
they can be sufficiently described as inducing explainability directly
and indirectly by penalizing: (1) the contrastive divergence,
\(\text{div}(\cdot)\), between mature counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and Mordatch
2020}). The trade-off between the components can be governed by
adjusting the strengths of the penalties \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed sample from
target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\). Conditional on the
target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes a mature
counterfactual for a randomly sampled factual from a non-target class
generated with a gradient-based CE generator for up to \(T\) iterations.
Mature counterfactuals are ones that have either reached convergence
wrt. the decision threshold \(\tau\) or exhausted \(T\).

@@

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation}

Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2013}) who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we choose
to work with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}) who consider
misclassification as the key distinguishing feature of AE. One of the
key observations in this work is that we can leverage CEs during
training and get adversarial examples essentially for free.

@@

\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation}

and introduce a novel divergence metric,

\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation}

where \(\mathbf{X}^\prime\) denotes a set of multiple counterfactuals
and \(\text{MMD}(\cdot)\) is an unbiased estimate of the squared
population maximum mean discrepancy
(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).

@@

\textbf{\emph{Interventions on features may impact fairness
downstream.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, counterfactual training could result in an unfairly
assigned burden of recourse (e.g.,
\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
threatening the equality of opportunity in the system
(\citeproc{ref-bell2024fairness}{Bell et al. 2024}) and potentially
reinforcing social segregation (\citeproc{ref-gao2023impact}{Gao and
Lakkaraju 2023}). Still, as the referenced publications indicate, such
phenomena are not specific to CT; all types of ADM solutions without
strong external protections have been recognized to promote harmful
power dynamics (\citeproc{ref-maas2023machine}{Maas 2023}).

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

@@

Overall, we find that Counterfactual Training (CT) achieves it key
objectives consistently across all hyperparameter settings and also
broadly across datasets: plausibility is improved by up to
\textasciitilde60 percent (\%) for the \emph{Circles} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-circles}), \textasciitilde25-30\%
for the \emph{Moons} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-moons}) and \textasciitilde10-20\%
for the \emph{Linearly Separable} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
average costs of faithful counterfactuals are reduced in many cases by
around \textasciitilde20-25\% for \emph{Circles} (e.g.
Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
\textasciitilde50\% for \emph{Moons} (e.g.
Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
Separable} data, costs are generally increased although typically by
less than 10\% (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
which reflects a common tradeoff between costs and plausibility
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Package Version (Reproducibility)}, colframe=quarto-callout-warning-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

>>>>

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{counterfactual training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: counterfactual training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
  follow the standard ML convention, where ``degrees of freedom'' refer
  to the number of parameters estimated from data.} Indeed, recent work
on the so-called ``lottery ticket hypothesis'' suggests that modern
neural networks can be pruned by up to 90\% while preserving their
predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
Carbin 2019}) and generalizability
(\citeproc{ref-morcos2019success}{Morcos et al. 2019}). Similarly, Zhang
et al. (\citeproc{ref-zhang2021understanding}{2021}) showed that
state-of-the-art neural networks are so expressive that they can fit
randomly labeled data. Thus, looking at the predictive performance
alone, the solutions may seem to provide compelling explanations for the
data, when in fact they are based on purely associative, semantically
meaningless patterns. This poses two related challenges. Firstly, there
is no dependable way to verify if such complex representations
correspond to meaningful and plausible explanations. Secondly, even if
we could resolve the first challenge, it remains undecided how to ensure
that models can \emph{only} learn valuable explanations.

@@

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data
(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
mitigates the aforementioned risk of relying on a single locally optimal
representations that corresponds to semantically meaningless
explanations for the data. Previous work by Schut et al.
(\citeproc{ref-schut2021generating}{2021}) similarly found that
generating plausible (``interpretable'') counterfactual explanations is
almost trivial for deep ensembles that have also undergone adversarial
training. The case for JEMs is even clearer: they involve a hybrid
objective that induces both high predictive performance and generative
capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
is closely related to the idea of aligning models with plausible
explanations and has inspired our proposed CT objective, as we explain
in Section~\ref{sec-method}.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit goal of aligning representations with plausible explanations
that comply with user requirements. In the context of CEs, plausibility
has broadly been defined as the degree to which counterfactuals comply
with the underlying data-generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CEs to provide
algorithmic recourse (AR) to individuals (negatively) affected by opaque
models. For AR recommendations to be actionable, they need to not only
result in plausible counterfactuals but also be attainable. A plausible
CE for a rejected 20-year-old loan applicant, for example, might reveal
that their application would have been accepted, if only they were 20
years older. Ignoring all other features, this would comply with the
definition of plausibility if 40-year-old individuals were in fact more
credit-worthy on average than young adults. But of course this CE does
not qualify for providing actionable recourse to the applicant since
\emph{age} is not a (directly) mutable feature. CT aims to improve model
explainability by aligning models with counterfactuals that meet both
desiderata: plausibility and actionability. Formally, we define
explainability as follows:

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms in Equation~\ref{eq-obj} are explained in detail
below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: (1) the
contrastive divergence, \(\text{div}(\cdot)\), between mature
counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and, (2) the adversarial loss, \(\text{advloss}(.)\),
with respect to nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\).
Finally, \(\text{ridge}(\cdot)\) denotes a Ridge penalty
(\(\ell_2\)-norm) that regularizes the magnitude of the energy terms
involved in \(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and
Mordatch 2020}). The trade-off between the components can be governed by
adjusting the strengths of the penalties \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function. We set
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Mature counterfactuals are ones
that have either reached convergence wrt. the decision threshold
\(\tau\) or exhausted \(T\).

@@

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2013}) who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we choose
to work with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}) who consider
misclassification as the key distinguishing feature of AE. One of the
key observations in this work is that we can leverage CEs during
training and get adversarial examples essentially for free.

@@

\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence metric,

\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a set of multiple
counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
squared population maximum mean discrepancy
(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).

@@

\textbf{\emph{Interventions on features may impact fairness
downstream.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, CT could result in an unfairly assigned burden of
recourse (e.g., \citeproc{ref-sharma2020certifai}{Sharma, Henderson, and
Ghosh 2020}), threatening the equality of opportunity in the system
(\citeproc{ref-bell2024fairness}{Bell et al. 2024}) and potentially
reinforcing social segregation (\citeproc{ref-gao2023impact}{Gao and
Lakkaraju 2023}). Still, as the referenced publications indicate, such
phenomena are not specific to CT; all types of ADM solutions without
strong external protections have been recognized to promote harmful
power dynamics (\citeproc{ref-maas2023machine}{Maas 2023}).

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note}, arc=.35mm, coltitle=black, left=2mm]

@@

Overall, we find that counterfactual training (CT) achieves it key
objectives consistently across all hyperparameter settings and also
broadly across datasets: plausibility is improved by up to
\textasciitilde60 percent (\%) for the \emph{Circles} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-circles}), \textasciitilde25-30\%
for the \emph{Moons} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-moons}) and \textasciitilde10-20\%
for the \emph{Linearly Separable} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
average costs of faithful counterfactuals are reduced in many cases by
around \textasciitilde20-25\% for \emph{Circles} (e.g.
Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
\textasciitilde50\% for \emph{Moons} (e.g.
Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
Separable} data, costs are generally increased although typically by
less than 10\% (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
which reflects a common tradeoff between costs and plausibility
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, colbacktitle=quarto-callout-warning-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Package Version (Reproducibility)}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
